---
title: "README"
author: "Rafa Zhong"
date: "2025-07-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---

# What does this document intend to do?

This document explains the steps taken to clean and summarize the Human Activity Recognition dataset.

---
title: "Codebook"
author: "Rafa Zhong"
date: "2025-07-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---

# What is this README about?

This codebook describes the variables and transformations performed in the "Getting and Cleaning Data" course project.

---

## Source Data

The original data is the [UCI Human Activity Recognition Using Smartphones Dataset](https://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones). It includes accelerometer and gyroscope measurements collected from Samsung Galaxy S smartphones worn by 30 subjects while performing 6 activities.

The original dataset is separated into training and testing sets, with measurements for multiple features recorded in files under `train/` and `test/` directories.

This codebook focuses on the intermediate merged data (`merged_data`) and the final tidy summary dataset (`tidy_dataset.txt`).

---

## Processing Steps for `merged_data`
1. The following files were read and combined:
   - `X_train.txt` and `X_test.txt`: raw measurement data
   - `y_train.txt` and `y_test.txt`: activity codes
   - `subject_train.txt` and `subject_test.txt`: subject IDs

2. The training and test data were merged using `rbind()`.

3. Variable names were assigned using `features.txt`.

4. Columns were labeled:
   - `subject`: identifies each participant
   - `activity`: integer code representing the activity performed (to be later mapped to names)
   - Remaining columns are measurements of signals from accelerometer and gyroscope sensors.

5. The resulting dataset (`merged_data`) contains:
   - 10299 rows
   - 563 columns (561 measurements + subject + activity)

6. Activities were recorded as integer codes (1 to 6), corresponding to:
   - `WALKING`, `WALKING_UPSTAIRS`, `WALKING_DOWNSTAIRS`, `SITTING`, `STANDING`, `LAYING`

---

## Variable Name Transformation Logic

To make the variable names more descriptive and human-readable, the original variable names from `features.txt` were transformed using the following rules:

- Prefix `"t"` was replaced with `"TimeDomain."`, indicating the signal is from the time domain.
- Prefix `"f"` was replaced with `"FrequencyDomain."`, indicating the signal is from the frequency domain.
- `"Acc"` was replaced with `"Accelerometer"` (sensor type).
- `"Gyro"` was replaced with `"Gyroscope"` (sensor type).
- `"BodyBody"` was corrected to `"Body"`.
- `"Body"` was changed to `"Body."` for consistency in hierarchy.
- `"Gravity"` was changed to `"Gravity."`.
- `"Jerk"` was replaced with `"JerkSignal"` (indicating signal derivative).
- `"Mag"` was replaced with `"Magnitude"` (vector norm).
- `-mean()` was replaced with `.Mean`
- `-std()` was replaced with `.STD`
- `-X`, `-Y`, `-Z` were replaced with `.XAxis`, `.YAxis`, `.ZAxis`
- Remaining parentheses and hyphens were removed.

The result is a set of variables that follow the hierarchical structure:

> `TimeDomain.Body.Accelerometer.Mean.XAxis`  
> = Mean value of body acceleration signal in the time domain, along the X axis.

---

## Variable Descriptions

This dataset contains sensor signals collected from a smartphone’s **accelerometer** and **gyroscope** worn by 30 human subjects while performing daily activities like walking, sitting, or laying. The measurements are from two signal domains:

- **TimeDomain.***: raw or filtered signals in the time domain
- **FrequencyDomain.***: transformed signals in the frequency domain (via Fast Fourier Transform)

Each signal can capture:
- **Body acceleration** or **gravity acceleration**
- **Linear acceleration** or **Jerk signals** (i.e., the rate of change of acceleration)
- **Angular velocity** (from gyroscope)
- **Magnitude** (i.e., the combined strength of a 3D signal vector)

These measurements are computed along **three axes (X, Y, Z)**, or as a **magnitude value** (without direction), and are summarized using:
- `.Mean`: average value
- `.STD`: standard deviation (i.e., how much variation there is)

Each variable in the dataset describes the **average** value of one of these measurements for a specific subject performing a specific activity.

---

## Units of Measurement

- Accelerometer measurements are in units of gravity (g).
- Gyroscope measurements are in radians per second (rad/s).
- All variables are normalized and bounded within [-1, 1].

---

## Examples of Variable Names and Their Meanings

| Variable Name | Description |
|---------------|-------------|
| `subject` | ID number of the subject (ranges from 1 to 30) |
| `activity` | The activity performed (e.g., WALKING, SITTING) |
| `TimeDomain.Body.Accelerometer.Mean.XAxis` | Average acceleration (in time domain) from the body sensor, along the X axis |
| `TimeDomain.Gravity.Accelerometer.STD.ZAxis` | Standard deviation of gravity acceleration (in time domain) along the Z axis |
| `TimeDomain.Body.Gyroscope.JerkSignal.Mean.YAxis` | Average rate of change of angular velocity (jerk) from body gyroscope, Y axis |
| `FrequencyDomain.Body.Accelerometer.Magnitude.Mean` | Average of the magnitude (strength) of the frequency-domain body acceleration signal |
| `FrequencyDomain.Body.Gyroscope.Magnitude.STD` | Standard deviation of the magnitude of body gyroscope in frequency domain |

---

## Structure of `tidy_dataset.txt`

- 180 rows (30 subjects × 6 activities)
- 68 columns:
  - `subject`: integer from 1 to 30
  - `activity`: one of six activities (e.g., `WALKING`, `LAYING`)
  - 66 measurement variables, all of which represent the average of a feature for each subject/activity pair. 

---

## Variable Descriptions

Each variable represents the **average** of a specific sensor measurement. For example:

- `Average.TimeDomain.Body.Accelerometer.Mean.XAxis`: Mean of time-domain body acceleration signal along X axis (unit: g)
- `Average.TimeDomain.Gravity.Accelerometer.STD.YAxis`: Standard deviation of gravity acceleration along Y axis
- `Average.FrequencyDomain.Body.Gyroscope.Magnitude.Mean`: Mean of frequency-domain body gyroscope magnitude

Variables follow the naming pattern:

> `Average.<Domain>.<Source>.<Sensor>.<Statistic>.<Axis (if any)>`

Where:
- Domain: TimeDomain or FrequencyDomain
- Source: Body or Gravity
- Sensor: Accelerometer or Gyroscope (possibly with Jerk or Magnitude)
- Statistic: Mean or STD
- Axis: XAxis, YAxis, ZAxis (optional for Magnitude)

